Predict Disease Outcome by Community Structures

### 2016-08-12 code/HumanCdGF_microbiota_predict.Rmd
compare day 0 comm to cfu

---
title: "Predictions From Day 0 Community Structures"
author: "Nicholas Lesniak"
date: "July 13, 2016"
output: html_document
---

@@

importance_df <- data.frame(matrix(nrow = length(OTU_list), 
                                   ncol = max(meta_file$day, na.rm = T)),
                            row.names = OTU_list)
rsq_df <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
                         row.names = 'RSQ')
prediction_df <- data.frame(matrix(nrow = 0, ncol = 4))

for (i in 1:10){
  predict_all_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day' , 'log_cfu') ],
                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
  predict_df <- predict_all_df[,! colnames(predict_all_df) %in% 'day']
  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
  rsq_df[i] <- RF$rsq[length(RF$rsq)]
  # sort the features by their effect on % increase in the standard error
  importance_df[ , i] <- importance(RF)[,1]
  colnames(importance_df)[i] <- paste('day', i, 'importance', sep = '_')
# fit the model back on the datar and tack it to the end of the counts file
  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
                              RF$predicted)
  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
  prediction_df <- rbind(prediction_df, random_forest_output)  
}

@@

n_important <- 20
median_importance <- names(head(sort((apply(importance_df,1, median)), decreasing = T), n_important))
med_import_df <- melt(cbind(rownames(importance_df), importance_df)[median_importance, ])
colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
        geom_point() + scale_x_discrete(labels = c(1:10)) +
        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
        x= 'Day', y = '% Increase in MSE') + theme_bw()
        coord_flip()

@@

##Can we predict the endpoint C. difficile CFU with the community structure at Day 0?
But endpoint has data from Day 2,3,10

Might be misleading since contains CFU from different data points

@@

RF_CFU_day_euth <- randomForest(log_cfu ~ ., data = CFU_euth[!is.na(CFU_euth$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
rsq_day_euth <- RF_CFU_day_euth$rsq[length(RF_CFU_day_euth$rsq)]
# sort the features by their effect on % increase in the standard error
importance_day_euth <- importance(RF_CFU_day_euth)[,1]
# fit the model back on the datar and tack it to the end of the counts file
rf_predict_cfu_output <- merge(meta_file[meta_file$day==meta_file$Euthanized, c('Mouse_ID', 'day', 'log_cfu') ],
                              as.data.frame(predict(RF_CFU_day_euth, rel_abund_stool)), 
                              by='Mouse_ID', by.y='row.names')
colnames(rf_predict_cfu_output)[3:4] <- c('observed', 'predicted')

str(RF_CFU_day_euth)
TEST <- cbind(CFU_euth[!is.na(CFU_euth$log_cfu), 'log_cfu'], data.frame(RF_CFU_day_euth$predicted))
colnames(TEST) <- c('observed', 'predicted')
ggplot(data = TEST, aes(x=observed, y=predicted)) + geom_point()

ggplot(data = rf_predict_cfu_output, aes(x=observed, y=predicted)) + geom_point() + 
        labs(title = 'Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community',
        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw()
        ggtitle(expression(atop("Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community",
                                atop(paste('% Var Explained = ', rsq_day_euth), ""))))

@@

ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
        geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU)) +
        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
        x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()

```

@@

##Can we predict the Day 10 C. difficile CFU with the community structure at Day 0?

```{r}
Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
Day_10_cdiff_cfu_rf$plot
```

% Variance explained `r Day_10_cdiff_cfu_rf$rf$rsq[length(Day_10_cdiff_cfu_rf$rf$rsq)]*100`

Why day 10? What is significant about day 10?

@@

##Does Day 0 community structure predict the each day C. difficile CFU at a similar?

##Can we predict the Day 1 C. difficile CFU with the community structure at Day 0?

```{r}
#rf to predict 
Day_1_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_1_cdiff_cfu')
Day_1_cdiff_cfu_rf$plot
```

% Variance explained `r Day_1_cdiff_cfu_rf$rf$rsq[length(Day_1_cdiff_cfu_rf$rf$rsq)]*100`

# Day 2 C. difficile CFU
```{r}
Day_2_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_2_cdiff_cfu')
Day_2_cdiff_cfu_rf$plot
```

% Variance explained `r Day_2_cdiff_cfu_rf$rf$rsq[length(Day_2_cdiff_cfu_rf$rf$rsq)]*100`

# Day 3 C. difficile CFU
```{r}
Day_3_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_3_cdiff_cfu')
Day_3_cdiff_cfu_rf$plot
```

% Variance explained `r Day_3_cdiff_cfu_rf$rf$rsq[length(Day_3_cdiff_cfu_rf$rf$rsq)]*100`

# Day 4 C. difficile CFU
```{r}
Day_4_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_4_cdiff_cfu')
Day_4_cdiff_cfu_rf$plot
```

% Variance explained `r Day_4_cdiff_cfu_rf$rf$rsq[length(Day_4_cdiff_cfu_rf$rf$rsq)]*100`

# Day 5 C. difficile CFU
```{r}
Day_5_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_5_cdiff_cfu')
Day_5_cdiff_cfu_rf$plot
```

% Variance explained `r Day_5_cdiff_cfu_rf$rf$rsq[length(Day_5_cdiff_cfu_rf$rf$rsq)]*100`

# Day 6 C. difficile CFU
```{r}
Day_6_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_6_cdiff_cfu')
Day_6_cdiff_cfu_rf$plot
```

% Variance explained `r Day_6_cdiff_cfu_rf$rf$rsq[length(Day_6_cdiff_cfu_rf$rf$rsq)]*100`

# Day 7 C. difficile CFU
```{r}
Day_7_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_7_cdiff_cfu')
Day_7_cdiff_cfu_rf$plot
```

% Variance explained `r Day_7_cdiff_cfu_rf$rf$rsq[length(Day_7_cdiff_cfu_rf$rf$rsq)]*100`

# Day 8 C. difficile CFU
```{r}
Day_8_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_8_cdiff_cfu')
Day_8_cdiff_cfu_rf$plot
```

% Variance explained `r Day_8_cdiff_cfu_rf$rf$rsq[length(Day_8_cdiff_cfu_rf$rf$rsq)]*100`

# Day 9 C. difficile CFU
```{r}
Day_9_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_9_cdiff_cfu')
Day_9_cdiff_cfu_rf$plot
```

% Variance explained `r Day_9_cdiff_cfu_rf$rf$rsq[length(Day_9_cdiff_cfu_rf$rf$rsq)]*100`

# Day 10 C. difficile CFU
```{r}
Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
Day_10_cdiff_cfu_rf$plot
```

% Variance explained `r Day_10_cdiff_cfu_rf$rf$rsq[length(Day_10_cdiff_cfu_rf$rf$rsq)]*100`

Predicting CFU from day 1 to 10 based on Day 0 community

There seems a flucuation in the % variance explained, which seems to coincide with the fluctuations in cdiff cfu

## seed greatly affects variance, likely due to data split, sample size might be too small to make reliable predictions

```{r knitr::kable, warning=F, message=F}
var_by_day <- data.frame(Day=c(1:10),Percent_Variance_Explained=NA)
cdiff_cfu <- vector()
for (a in 1:10){
          var_by_day$Percent_Variance_Explained[a] <- eval(parse(text = paste('Day_', a, '_cdiff_cfu_rf$rf$rsq[length(Day_', a, '_cdiff_cfu_rf$rf$rsq)]*100', sep='')))
          cdiff_cfu <- c(cdiff_cfu,paste('Day_', a, '_cdiff_cfu', sep=''))
     }

knitr::kable(var_by_day)
plot(var_by_day, type='n', xlab='Day',ylab='Percent Var Explained')
lines(var_by_day$Day, var_by_day$Percent_Variance_Explained)

cfu_plot <- cbind(Mouse = paste(Predict_df$human_source,Predict_df$mouse_eartag, sep='_'),Predict_df[, cdiff_cfu])
names(cfu_plot) <- c('Mouse', c(1:10))
cfu_plot <- reshape2::melt(cfu_plot, by='Mouse')

ggplot(data=cfu_plot, aes(x=variable, y=value, group=Mouse)) + geom_line(aes(color=Mouse)) + geom_point() + scale_y_log10() +
     theme_classic() + theme(legend.position='none') + labs(title='C. Difficile CFU by mouse',x='Day',y='CFU')
```

***
***

##Can we predict day 1 weight change with the community structure at Day 0?
```{r}

```

##Can we predict day 10 weight change with the community structure at Day 0?
```{r}

```

##Can we predict ___________ with the community structure at Day 0?
```{r}

```

>>>>

---
title: "Predict Disease Outcome by Community Structures"
author: "Nicholas Lesniak"
date: "July 13, 2016"
output: html_document
---

@@

importance_df <- data.frame(matrix(nrow = length(OTU_list), 
                                   ncol = max(meta_file$day, na.rm = T)),
                            row.names = OTU_list)
rsq_df <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
                         row.names = 'RSQ')
mse_df <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
                         row.names = 'MSE')
prediction_df <- data.frame(matrix(nrow = 0, ncol = 4))

for (i in 1:10){
  predict_all_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day' , 'log_cfu') ],
                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
  predict_df <- predict_all_df[,! colnames(predict_all_df) %in% 'day']
  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
  rsq_df[i] <- RF$rsq[length(RF$rsq)]
  mse_df[i] <- RF$mse[length(RF$mse)]

  # sort the features by their effect on % increase in the standard error
  importance_df[ , i] <- importance(RF)[,1]
  colnames(importance_df)[i] <- paste('day', i, 'importance', sep = '_')
# fit the model back on the datar and tack it to the end of the counts file
  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
                              RF$predicted)
  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
  prediction_df <- rbind(prediction_df, random_forest_output)  
}

@@

mse_df <- t(mse_df)
mse_df <- as.data.frame(cbind(mse_df, day = c(1:10)))
ggplot(data = mse_df, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
        x= 'Day', y = 'Variance Explained')

n_important <- 20
median_importance <- names(head(sort((apply(importance_df,1, median)), decreasing = T), n_important))
med_import_df <- melt(cbind(rownames(importance_df), importance_df)[median_importance, ])
colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
        geom_point() + scale_x_discrete(labels = c(1:10)) +
        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
        x= 'Day', y = '% Increase in MSE') + theme_bw()

@@

##Can we predict the endpoint C. difficile CFU with the community structure at Day 0?
Endpoint has data from Day 2,3,10 - might be misleading since contains CFU from different data points

@@

RF_CFU_day_euth <- randomForest(log_cfu ~ ., data = CFU_euth[!is.na(CFU_euth$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
rsq_day_euth <- round(RF_CFU_day_euth$rsq[length(RF_CFU_day_euth$rsq)], 2)
mse_day_euth <- round(RF_CFU_day_euth$mse[length(RF_CFU_day_euth$mse)], 2)
# sort the features by their effect on % increase in the standard error
importance_day_euth <- importance(RF_CFU_day_euth)[,1]
# fit the model back on the datar and tack it to the end of the counts file
rf_predict_cfu_output <- cbind(CFU_euth[!is.na(CFU_euth$log_cfu), 'log_cfu'], data.frame(RF_CFU_day_euth$predicted))
colnames(rf_predict_cfu_output) <- c('observed', 'predicted')

ggplot(data = rf_predict_cfu_output, aes(x=observed, y=predicted)) + geom_point() + 
        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
        ggtitle("Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community") +
        annotate('text', x=0.15*(max(rf_predict_cfu_output$observed)), y=0.95*(max(rf_predict_cfu_output$predicted)), 
                 label=paste('R^2 = ', rsq_day_euth,'\nMSE = ', mse_day_euth, sep = ''))

@@

ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
        geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU)) +
        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community',
        x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
```

@@

@@

##Does Day Community structure predict the each day C. difficile CFU?

RF all mice/days together
RF cfu vs community by day
